Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-12-13 analysis/paper/paper.Rmd

Ice-associated seals rely on sea ice for a variety of activities, including
breeding, molting, pupping, and resting. In the Arctic, many of these activities
occur in spring (April-June) as sea ice begins to melt and retreat northward.
Rapid acceleration of climate change in Arctic ecosystems is therefore of
concern as the quantity and quality of suitable habitat is forecast to decrease.
In this paper, we use data collected from satellite-linked bio-loggers deployed
between 2005 and 2020 to investigate the seasonal timing and environmental
factors affecting sea ice use by seals (specifically, bearded, ribbon, and
spotted seals) in the Bering and Chukchi seas. In addition to providing baseline
data on phenology, these data also allow us to quantify 'availability', which is
needed to accurately estimate abundance from aerial survey counts of seals
basking on ice (i.e., to correct for the proportion of animals that are in the
water while surveys are conducted). Using generalized linear mixed pseudo-models
to properly account for temporal autocorrelation, we fit models with a variety
of covariates (e.g., day-of-year, solar hour, age-sex class, wind speed,
barometric pressure, temperature, precipitation) to examine their ability to
explain variation in hourly haul-out records. We found evidence for strong
diurnal and within-season patterns in haul-out behavior, as well as strong
weather effects (particularly wind and temperature). In general, seals were more
likely to rest on ice in the middle of the day and when wind speed was low and
temperatures were higher. Haul-out propensity increased through March and April,
peaking in May and early June before declining again. The timing and frequency
of haul-out events also varied based on species and age-sex class. For ribbon
and spotted seals, models with year effects were highly supported, indicating
that the timing and magnitude of haul-out behavior varied among years. However,
we did not find evidence that haul-out timing was linked to annual sea ice
extent. Our analysis emphasizes the importance of accounting for variation in
haul-out behavior when interpreting the number of seals counted in aerial
surveys.

@@

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals
rest out of the water almost exclusively on sea ice and are mostly pelagic
outside the spring pupping, breeding, and molting season [@burns1981g]. While
spotted and bearded seals rest on beaches during ice-free months, they strongly
prefer sea ice as a resting platform during the spring and early summer
[@burns2009]. The remoteness of the Bering and Chukchi seas mean direct
scientific observation of seal behavior is impractical. Bio-logging devices are,
thus, especially useful tools for collecting key information on movement and
haul-out behavior for these species.

Bio-logging records of time spent out of the water provide valuable data for
identifying covariates that explain variation in haul-out behavior. For
instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
bio-logging data to demonstrate greater haul-out propensity in juvenile
crabeater seals (*Lobodon carcinophaga*) than adults, with highest probabilities
in February and at times close to solar noon. In the Arctic, Von Duyke et al.
[-@vonduyke2020] used seal-borne, satellite-linked loggers to corroborate
seasonal changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly et al. [-@kelly2010] using labor-intensive VHF
radio tags and direct observation. Bengtson et al. [-@bengtson2005a] documented
a higher propensity for ringed seal basking near solar noon, as did Ver Hoef et
al. [-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using
much larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion
of time bearded seals spent hauled out progressively increased through spring
and summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
increased gradually starting in March and peaking in May and June for bearded,
ribbon, and spotted seals.

While bio-logging of haul-out patterns clearly has utility for description of
ice-associated seals' natural history and ecology, some studies have focused, as
we do here, specifically on quantifying the probability of seals being hauled
out as a means of accounting for seals in the water, and therefore missed,
during aerial surveys. These studies typically use logistic regression-style
analyses to estimate the time-specific probability of being hauled out base on
'wet/dry' data relayed by bio-loggers. In these models, haul-out probabilities
were expressed as a function of predictive covariates, such as time of day,
day-of-year, sex, age class, and environmental conditions (e.g., [@reder2003],
[@bengtson2004], [@bengtson2005a], [@udevitz2009], [@verhoef2014a], and
[@southwell2008]). However, sample sizes have often been low and insufficient to
permit strong inference about life history and/or seasonal variation in haul-out
probabilities. For instance, Bengtson and Cameron's [-@bengtson2004] study
included 5 adult and 2 juvenile crabeater seals, while Bengtson et al.'s
[-@bengtson2005a] study was based on 6 telemetered ringed seals in the Chukchi
Sea. These studies were often further limited by logistical constraints on
tagging fieldwork and the attachment duration or operational life of satellite
tags. In this study, we address some of these limitations by deploying small
bio-loggers designed for attachment on a portion of the study individuals. These
devices are designed to collect data through the molt period (when those adhered
to the hair would fall off) and, in some situations, provide multiple years of
data.

Ultimately, knowledge of trends in phenology and abundance (or life history
surrogates such as survival and recruitment) will be necessary to make credible
quantitative predictions about the effects of the climate crisis on the
abundance and distribution of Arctic seal populations. Before we can construct a
trend, however, we first require a baseline. Several studies have contributed
estimates of the distribution and abundance of ice-associated seal species in
the Arctic using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
[@verhoef2014a]). Such studies were conducted over very large areas and
estimation of absolute abundance required making inference about numerous
nuisance processes affecting the observation of seals on ice. This included
availability (only seals basking on ice were available to be counted), detection
probability (observers or automated detection systems may have missed some seals
on ice), species misclassification, and possible disturbance of seals by
aircraft [@conn2014a;
@verhoef2014a].

In this study, we used data from 16 years of bio-logger records in a
comprehensive investigation of haul-out behavior by 3 species of ice-associated
seals in the Bering and Chukchi seas. We focused our effort on bearded, ribbon,
and spotted seals and did not include ringed seals in the analysis. Ringed seals
are unique in their use of snow lairs in the spring months, and they warrant a
separate study that includes additional environmental covariates, such as the
onset of snow melt and snow depth. Our goals are threefold. First, we wish to
establish baseline estimates for the chronology of haul-out behavior in the
critical spring season for each species across different age and sex classes.
Second, we wish to refine estimates of haul-out availability corrections for
aerial surveys in order to improve estimates of seal abundance. Previously
estimated availability correction factors (e.g., [@bengtson2005a], [@conn2014a],
and [@verhoef2014a]) accounted for variables such as the time-of-day and
day-of-year, but did not investigate meteorological variables that have been
shown to influence haul-out behavior of walruses [@reder2003; @udevitz2009].
Third, we explored the annual variability in haul-out timing and possible
linkage to dramatic changes in the availability and extent of seasonal sea ice
between 2005 and 2020. Our work extends the scope of previous haul-out analyses,
includes the influence of meteorological variability, and investigates the
potential impact of a changing icescape on the behavior of these species.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
bio-loggers deployed on bearded, ribbon, and spotted seals was filtered to
include only records from 1 March to 15 July between `r min(lubridate::year(analysis_data$haulout_dt))` 
and `r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
annual molt, which limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while
the seal is at the surface and can slightly bias the percent-dry values recorded
by the device. For this study, in cases where both tag types were deployed,
percent dry observations from the flipper tag were preferred. Sex and age class
(non-dependent *young-of-the-year*, sexually immature *subadults*, and mature
*adults*) were determined at the time of deployment by subjective judgement
based on length, and pelage characteristics for some species. Field
identification of age class can be inexact, particularly when discerning
subadults from adults. In the case of ribbon seals, pelage differences between
subadults and adults are distinguishable, with subadults often having less
defined ribbons. Bearded seal subadults will often have a spotted pattern in the
pelage that is not seen in adults. Spotted seal pelage is monomorphic and can
not be used to discern age class. Despite these challenges, we feel the age
classifications used in this analysis are useful categories in testing for
age-related effects on haul-out behavior. Seals determined to be less than one
year were classified as young-of-the-year. Subadults are those seals likely
greater than one year of age but not yet sexually mature. Adults are likely
sexually mature and older than approximately four years. For those tags deployed
on young-of-the-year and transmitting into the next year, the age class was
advanced to subadult on 1 March of the following year. Table
\@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

```{r tblDeploy, include=TRUE, tab.id = "tblDeploy", tab.cap = "Summary of bio-logger deployments across seal species and age classification from 1 March to 15 July. Total seal hours represents the sum of available data across all seals."}

@@

Tags that fall off due to molt or mortality and remain on ice or land can still
send data to satellites, and we did not want to include data from these tags in
our analyses. As such, end times of each deployment were identified by experts
who examined several simultaneous data streams (e.g., tag locations and dive
behavior) to determine when tags ceased providing data consistent with seal
behavior. Sensor readings outside of these start and end times were discarded
prior to analysis. After approximately 9 months, 7 devices deployed on the rear
flipper of bearded seals reported implausible sensor data. All data after the
first instance of unrealistic values were censored from this analysis. Figure
\@ref(fig:dataCal) shows a calendar-like distribution of these data across the
study season for each species. Observations for ribbon and spotted seals are
concentrated in the months of May and June due to the timing of deployment
(April and May) and the timing of molt (May and June). During molt, seals (and
their attached bio-loggers) spend more time out of the water and more data is
transmitted. Molt timing also impacts when many deployments end as any tags
adhered to the hair will fall off. Bearded seal deployments are more evenly
spread with fewer observations in May and June. The majority of bearded seal
deployments start later in the summer and by May tags have either fallen off or
the batteries are depleted.

```{r seal-hours-calendar, include=TRUE, fig.height = 4.75, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species. The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, a majority of deployments started in April and May. Data were collated across all years between 2005 and 2020."}
ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,4.3), breaks = c(100,500,1000,1500)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(13,18,23,27),
                     labels = c("April", "May", "June", "July")) +
  scale_y_discrete(limits = rev) +
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(
        #axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(x = "", y = "", color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("The number of haul-out behavior observations varies from March through July",
          subtitle = "data presented in seal-hours collated across years (2005-2020)")
```

Haul-out behavior data was recorded and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and a percent of the
hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). Each data
transmission message consisted of a complete 24-hour record. Tags were,
generally, deployed and programmed in a manner to maximize data transmission
during the spring pupping and molting period, however, not all days during the
deployment were successfully transmitted. This is due to a variety of factors
including satellite coverage, tag availability (i.e. tags mounted to the rear
flipper are not available to the surface while the seal is at sea), tag
performance, duty cycling, and extra-terrestrial interference.

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, and May. Empty areas of the plot within a month represent missing data that were not successfully transmitted from the tag."}
d <- analysis_data %>% 
  filter(speno == "HF2016_1002") %>% 
  dplyr::select(speno, haulout_dt, percent_dry) %>% 
  arrange(haulout_dt) %>% st_drop_geometry() %>% 
  mutate(year = lubridate::year(haulout_dt),
         month = lubridate::month(haulout_dt,label=TRUE),
         day = lubridate::day(haulout_dt),
         hour = lubridate::hour(haulout_dt)) %>% 
  filter(month != 'Jun')

@@

p <- p + facet_grid(year~month, scales = "free_x")
#p <- p + scale_y_continuous(trans = "reverse", expand = c(0,0))
p <- p + scale_x_continuous(breaks = c(5,15,25), expand = c(0,0)) +
  scale_y_continuous(breaks = c(4,12,20))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  theme(panel.spacing = unit(0,"line")) +
  ggtitle("Observation records are incomplete because not all data are transmitted",
          subtitle = "percent-dry data for a single ribbon seal across two years") +
  xlab("day of month") + ylab("hour (UTC)")
p
```

Of key interest in this study is the relationship between haul-out behavior and
a variety of environmental covariates that vary with space and time. We explored
the use of a movement model to predict locations at specific times. However, the
sparse nature of data from some tags, especially those mounted to the rear
flipper, resulted in poor modeling performance or convergence issues. For this
study, we calculated a weighted average daily location where the inverse of the
estimated Argos location error was used for the weight. Any days where haul-out
observations were present without any location data were filled from the last
calculated weighted average daily location, and any days where the location
intersected with land were removed from the dataset. Figure \@ref(fig:dataMap)
shows the spatial distribution of determined locations and available haul-out
behavior data for each species across the study area.

@@

```{r dataMap, fig.height = 3.75, include=TRUE, fig.id="dataMap", fig.cap="Spatial distribution of haul-out data during the months of March through July for each of the three species. Data were collated across all years between 2005 and 2020."}

@@

In addition to sex and age class, we gathered additional variables that might
help explain variation in haul-out probabilities. These included day-of-year
(for seasonal effects) and local solar hour (for diurnal effects). Solar hour
was calculated using the {solaR} package [@perpinan2012] within the R
statistical environment [@rcoreteam2021] using the daily locations. We also
compiled several meteorological variables shown to affect haul-out behavior in
other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017]. In particular, we
linked locations from satellite tags to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America.
Numerous weather variables are made available across the region 8 times daily.
For this study, NARR weather values were subset to the extent of our study area
over the Bering and Chukchi Seas at 3-hr intervals based on the native grid
resolution of 32 km (1024 sq. km). The following meteorological variables were
interpolated and assigned to seal locations using a bilinear method: 1) air
temperature at 2m above the earth's surface, 2) wind consisting of northerly and
easterly vector components converted to wind speed using the Euclidean norm, 3)
barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem
counterintuitive when modeling the haul-out behavior of seal species with such a
close association to sea ice; seals haul out in the presence of sea ice, and we
could assess the local concentration of sea ice during these events (see
[@olnes2020]). However, a major focus of this study is to develop models
applicable for aerial survey correction factors and using sea ice as a covariate
would almost certainly bias haul-out predictions towards those seals that are on
or near ice and therefore more likely to haul out. Since aerial surveys can only
detect seals on ice, abundance estimates would be missing a correction for those
seals that are away from ice (e.g., on foraging trips). Additionally, many of
the bio-logger deployments used in this study were single attachments to the
rear flipper of the seal and do not provide at-sea locations, limiting our
ability to fully evaluate habitat preferences related to sea-ice. It should be
noted that the aerial survey counts are modeled with sea-ice and other habitat
parameters. So, any final abundance estimates that combine models of survey
counts and our model for proportion of seals hauled out on ice do account for
the strong association between these species and sea-ice in the Bering and
Chukchi seas. Lastly, our study was limited to the spring season when seal
haul-out tendencies are strongly influenced by pupping, nursing, breeding, and
molt and these drivers are likely more influential than specific sea-ice
concentration. Crawford et al [-@crawford2019] compared haul-out probability
models for bearded seals and found those that only included season (and not
sea-ice concentration) were the most parsimonious. For these reasons, we have
elected not to use sea ice concentration as a predictor for haul-out
probability; studies to evaluate habitat selection in relation to sea-ice
parameters and other environmental covariates are in progress.

@@

Haul-out records for seals are often characterized by multiple hours spent
basking on ice alternating with long periods of foraging (Figure
\@ref(fig:examplePlot)). Commonly used statistical models for binary data (e.g.
logistic regression) assume independence among responses, an assumption that is
clearly violated if hourly responses are modeled. Any analysis that ignores
temporal dependence in responses will thus have overstated precision
[@betts2006].

>>>>

Ice-associated seals rely on sea ice for a variety of activities, including
breeding, molting, pupping, and resting. In the Arctic, many of these activities
occur in spring (April-June) as sea ice begins to melt and retreat northward.
Rapid acceleration of climate change in Arctic ecosystems is therefore of
concern as the quantity and quality of suitable habitat is forecast to decrease.
In this paper, we use data collected from satellite-linked bio-loggers deployed
between 2005 and 2020 to investigate the seasonal timing and environmental
factors affecting haul-out behavior sea ice by seals (specifically, bearded,
ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to
providing baseline data on phenology, these data also allow us to quantify
'availability', which is needed to accurately estimate abundance from aerial
survey counts of seals basking on ice (i.e., to correct for the proportion of
animals that are in the water while surveys are conducted). Using generalized
linear mixed pseudo-models to properly account for temporal autocorrelation, we
fit models with a variety of covariates (e.g., day-of-year, solar hour, age-sex
class, wind speed, barometric pressure, temperature, precipitation) to examine
their ability to explain variation in hourly haul-out records. We found evidence
for strong diurnal and within-season patterns in haul-out behavior, as well as
strong weather effects (particularly wind and temperature). In general, seals
were more likely to haul out on ice in the middle of the day and when wind speed
was low and temperatures were higher. Haul-out propensity increased through
March and April, peaking in May and early June before declining again. The
timing and frequency of haul-out events also varied based on species and age-sex
class. For ribbon and spotted seals, models with year effects were highly
supported, indicating that the timing and magnitude of haul-out behavior varied
among years. However, we did not find evidence that haul-out timing was linked
to annual sea ice extent. Our analysis emphasizes the importance of accounting
for variation in haul-out behavior when interpreting the number of seals counted
in aerial surveys.

@@

For ice-associated seals, haul-out behavior is tightly linked with sea ice.
Ribbon seals haul out of the water almost exclusively on sea ice and are mostly
pelagic outside the spring pupping, breeding, and molting season [@boveng2018a].
While spotted and bearded seals rest on beaches during ice-free months, they
strongly prefer sea ice as a resting platform during the spring and early summer
[@frost2018]. The remoteness of the Bering and Chukchi seas mean direct
scientific observation of seal behavior is impractical. Thus, bio-logging
devices are, thus, especially useful tools for collecting key information on
movement and haul-out behavior for these species.

Bio-logging records of time spent out of the water provide valuable data for
identifying covariates that explain variation in haul-out behavior. For
instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
bio-logging data to demonstrate greater haul-out propensity in juvenile
crabeater seals (*Lobodon carcinophaga*) than adults, with highest probabilities
in February and at times close to solar noon. In the Arctic, Von Duyke et al.
[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly et al. [-@kelly2010] using VHF radio tags and
direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
time bearded seals spent hauled out progressively increased through spring and
summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
increased gradually starting in March and peaking in May and June for bearded,
ribbon, and spotted seals.

While analyzing haul-out patterns clearly has utility for understanding natural
history and ecology, some studies have focused, as we do here, specifically on
quantifying the probability of seals being hauled out as a means of quantifying
seals in the water that are not counted during aerial surveys. These studies
typically use logistic regression-style analyses to estimate the time-specific
probability of being hauled out based on 'wet/dry' data relayed by bio-loggers.
In these models, haul-out probabilities were expressed as a function of
predictive covariates, such as time of day, day-of-year, sex, age class, and
environmental conditions (e.g., [@reder2003], [@bengtson2004], [@bengtson2005a],
[@udevitz2009], [@verhoef2014a], and [@southwell2008]). However, sample sizes
have often been insufficient to permit strong inference about life history
and/or seasonal variation in haul-out probabilities. For instance, Bengtson and
Cameron's [-@bengtson2004] study included 5 adult and 2 juvenile crabeater
seals, while Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed
seals in the Chukchi Sea. These studies were often further limited by logistical
constraints on tagging fieldwork and the attachment duration or operational life
of bio-loggers. In this study, we address some of these limitations by deploying
small bio-loggers designed for longer-term attachment on a subset of the study
individuals. These devices are designed to collect data through the molt period
(when those adhered to the hair would fall off) and, in some situations, provide
multiple years of data.

Ultimately, knowledge of trends in phenology and abundance (or life history
surrogates such as survival and recruitment) will be necessary to make credible
quantitative predictions about the effects of climate disruptions on the
abundance and distribution of Arctic seal populations. Before we can construct a
trend, however, we first require a baseline. Several studies have contributed
estimates of the distribution and abundance of ice-associated seal species in
the Arctic using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
[@verhoef2014a]). Such studies were conducted over very large areas and
estimation of absolute abundance required making inference about numerous issues
affecting the observation of seals on ice. These included availability (only
seals basking on ice were available to be counted), detection probability
(observers or automated detection systems may have missed some seals on ice),
species misclassification, and possible disturbance of seals by aircraft
[@conn2014a; @verhoef2014a].

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of 3 species of ice-associated seals in the Bering and Chukchi seas. We
focused our effort on bearded, ribbon, and spotted seals and did not include
ringed seals in the analysis. Ringed seals are unique in their use of snow lairs
in the spring months, and they warrant a separate study that includes additional
environmental covariates, such as the onset of snow melt and snow depth. Our
goals are threefold. First, we wish to establish baseline estimates for the
chronology of haul-out behavior in the critical spring season for each species
across different age and sex classes. Second, we wish to refine estimates of
haul-out availability corrections for aerial surveys in order to improve
estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not investigate
meteorological variables that have been shown to influence haul-out behavior of
walruses [@reder2003; @udevitz2009]. Third, we explored the annual variability
in haul-out timing and possible linkage to changes in the extent of seasonal sea
ice between 2005 and 2020. Our work extends the scope of previous haul-out
analyses, includes the influence of meteorological variability, and investigates
the potential impact of a changing icescape on the behavior of these species.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
bio-loggers deployed on bearded, ribbon, and spotted seals was filtered to
include only records from 1 March to 15 July between `r
min(lubridate::year(analysis_data$haulout_dt))` and `r
max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
annual molt, which limits the effective length of haul-out timelines.
Additionally, bio-loggers attached to the head or dorsal region are often dry
while the seal is at the surface and can slightly bias the percent-dry values
recorded by the device. For this study, in cases where both tag types were
deployed, percent dry observations from the flipper tag were preferred. Sex and
age class (non-dependent *young-of-the-year*, sexually immature *subadults*, and
mature *adults*) were determined at the time of deployment by varying
combinations of length, claw annuli, and pelage characteristics for some
species. Field identification of age class can be inexact, particularly when
discerning subadults from adults. In the case of ribbon seals, pelage
differences between subadults and adults are distinguishable, with subadults
often having less defined ribbons. Bearded seal subadults will often have a
spotted pattern in the pelage that is not seen in adults. Spotted seal pelage is
monomorphic and can not be used to discern age class. Despite these challenges,
we feel the age classifications used in this analysis are useful categories in
testing for age-related effects on haul-out behavior. Seals determined to be
less than one year were classified as young-of-the-year. Subadults are those
seals likely greater than one year of age but not yet sexually mature. Adults
are likely sexually mature and older than four years. For those
bio-loggers deployed on young-of-the-year and transmitting into the next year,
the age class was advanced to subadult on 1 March of the following year. Table
\@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

```{r tblDeploy, include=TRUE, tab.id="tblDeploy",tab.cap = "Summary of bio-logger data across seal species and age classification from 1 March to 15 July 2005-2020. Total seal hours represents the sum of available data across all seals."}

@@

Tags that fall off due to molt or mortality and remain on ice or land can still
send data to satellites, and we did not want to include data from these tags in
our analyses. As such, end times of each deployment were identified by examining
tag locations and dive behavior to determine when tags ceased providing data
consistent with seal behavior. Sensor readings outside of these start and end
times were discarded prior to analysis. After approximately 9 months, 7 devices
deployed on the rear flipper of bearded seals reported implausible sensor data.
All data after the first instance of unrealistic values were censored from this
analysis. Figure \@ref(fig:dataCal) shows a calendar-like distribution of all
haul-out data across the study season for each species. Observations for ribbon
and spotted seals are concentrated in the months of May and June due to the
timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the
water and more data is transmitted. Molt timing also impacts when many
deployments end as any tags adhered to the hair will fall off. Bearded seal
deployments are more evenly spread with fewer observations in May and June. The
majority of bearded seal deployments start later in the summer and by May tags
have either fallen off or the batteries are depleted.

```{r seal-hours-calendar, include=TRUE, fig.id = "dataCal", fig.height = 4.75, fig.cap = "Distribution of haul-out behavior data from 1 March to 15 July for each species. The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, many bio-logger deployments started in April and May. Data were collated across all years between 2005 and 2020."}

ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,4.3), breaks = c(100,500,1000,1500)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(13,18,23,27),
                     labels = c("April", "May", "June", "July")) +
  scale_y_discrete(limits = rev) +
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(
        #axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(x = "", y = "", color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("Density of haul-out behavior observations varies from March through July",
          subtitle = "data presented in seal-hours collated across years (2005-2020)")
```

Haul-out behavior data was recorded and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and a percent of the
hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). Each data
transmission message consisted of a complete 24-hour record. Bio-loggers were,
generally, deployed and programmed in a manner to maximize data transmission
during the spring pupping and molting period, however, not all days during a
deployment were successfully transmitted. This is due to a variety of factors
including satellite coverage, tag availability (i.e. tags mounted to the rear
flipper are not available to the surface while the seal is at sea), tag
performance, duty cycling, and extra-terrestrial interference.

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, and May. Areas of the plot indicated with an 'X' represent missing data that were not successfully transmitted from the tag."}

d <- analysis_data %>% 
  filter(speno == "HF2016_1002") %>% 
  dplyr::select(speno, haulout_dt, percent_dry) %>% 
  arrange(haulout_dt) %>% st_drop_geometry() %>% 
  mutate(year = lubridate::year(haulout_dt),
         month = lubridate::month(haulout_dt,label=TRUE),
         day = lubridate::day(haulout_dt),
         hour = lubridate::hour(haulout_dt)) %>% 
  filter(month != 'Jun')

d_deploy <- analysis_data %>% 
  sf::st_drop_geometry() %>% 
  filter(speno == "HF2016_1002") %>% 
  dplyr::select(speno,deploy_dt,haulout_dt,percent_dry) %>% 
  mutate(year = lubridate::year(deploy_dt),
         month = lubridate::month(deploy_dt, label = TRUE),
         day = lubridate::day(deploy_dt),
         hour = lubridate::hour(deploy_dt))

missing_hours <- tibble(haulout_dt = seq(min(d_deploy$haulout_dt),
              max(d_deploy$haulout_dt),by = "hour")) %>% 
  filter(!haulout_dt %in% d_deploy$haulout_dt) %>% 
  filter(lubridate::month(haulout_dt) %in% c(3,4,5)) %>% 
  mutate(percent_dry = 0) %>% 
  mutate(year = lubridate::year(haulout_dt),
         month = lubridate::month(haulout_dt,label=TRUE),
         day = lubridate::day(haulout_dt),
         hour = lubridate::hour(haulout_dt))

@@

p <- p + facet_grid(year~month, scales = "free_x")
p <- p + scale_x_continuous(breaks = c(5,15,25), expand = c(0,0)) +
  scale_y_continuous(breaks = c(4,12,20))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  theme(panel.spacing = unit(0,"line")) +
  ggtitle("Observation records are incomplete because not all data are transmitted",
          subtitle = "percent-dry data for a single ribbon seal across two years") +
  xlab("day of month") + ylab("hour (UTC)") 
p <- p + 
  geom_point(aes(x = day, y = hour), data = missing_hours, 
             color = "black", shape = 4, size=1)
p
```

Of key interest in this study is the relationship between haul-out behavior and
environmental covariates that vary with space and time. We explored
the use of a movement model to predict locations at specific times. However, the
sparse nature of data from some tags, especially those mounted to the rear
flipper, resulted in poor modeling performance or convergence issues. For this
study, we calculated a weighted average daily location where the inverse of the
estimated Argos location error was used for the weight. Any days where haul-out
observations were present without any location data were filled from the last
calculated weighted average daily location, and any days where the location
intersected with land were removed from the dataset. Figure \@ref(fig:dataMap)
shows the spatial distribution of determined locations and available haul-out
behavior data for each species across the study area.

@@

```{r dataMap, fig.height = 3.75, fig.id="dataMap", include=TRUE, fig.cap="Spatial distribution of haul-out data during the months of March through July 15 for each of the three species. Data were collated across all years between 2005 and 2020."}

@@

In addition to sex and age class, we analyzed variables that might
help explain variation in haul-out probabilities. These included day-of-year
(for seasonal effects) and local solar hour (for diurnal effects). Solar hour
was calculated using the {solaR} package [@perpinan2012] within the R
statistical environment [@rcoreteam2021] using the daily locations. We included
several meteorological variables shown to affect haul-out behavior in
other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017]. In particular, we
linked locations from our biologgers to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America.
Numerous weather variables are made available across the region 8 times daily.
For this study, NARR weather values were subset to the extent of our study area
over the Bering and Chukchi seas at 3-hr intervals based on the native grid
resolution of 32 km (1024 sq. km). The following meteorological variables were
interpolated and assigned to seal locations using a bilinear method: 1) air
temperature at 2m above the earth's surface, 2) wind consisting of northerly and
easterly vector components converted to wind speed using the Euclidean norm, 3)
barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem
counterintuitive when modeling the haul-out behavior of seal species with such a
close association to sea ice; seals haul out in the presence of sea ice, and we
could assess the local concentration of sea ice during these events (see
[@olnes2020]). However, a major focus of this study is to develop models
applicable for aerial survey correction factors and using sea ice as a covariate
would almost certainly bias haul-out predictions towards those seals that are on
or near ice and therefore more likely to haul out. Since aerial surveys can only
detect seals on ice, abundance estimates would be missing a correction for those
seals that are away from ice (e.g., on foraging trips). Additionally, many of
the bio-logger deployments consisted of a single device attached to the
rear flipper of the seal which do not provide at-sea locations, limiting our
ability to fully evaluate habitat preferences related to sea-ice. It should be
noted that the aerial survey counts are modeled with sea-ice and other habitat
parameters. So, any final abundance estimates that combine models of survey
counts and our model for proportion of seals hauled out on ice do account for
the strong association between these species and sea-ice in the Bering and
Chukchi seas. Lastly, our study was limited to the spring season when seal
haul-out tendencies are strongly influenced by pupping, nursing, breeding behavior, and
molt and these drivers are likely more influential than specific sea-ice
concentration. Crawford et al [-@crawford2019] compared haul-out probability
models for ringed seals and found those that only included season (and not
sea-ice concentration) were the most parsimonious. For these reasons, we have
elected not to use sea ice concentration as a predictor for haul-out
probability; studies to evaluate habitat selection in relation to sea-ice
parameters and other environmental covariates are in progress.

@@

Haul-out records for seals are often characterized by sequential hours spent
basking on ice alternating with long periods in the water (Figure
\@ref(fig:examplePlot)). Commonly used statistical models for binary data (e.g.
logistic regression) assume independence among responses, an assumption that is
clearly violated if hourly responses are modeled. Any analysis that ignores
temporal dependence in responses will thus have overstated precision
[@betts2006].
